---
title: "Data Visualisation with R"
author: "Datasoc - Daniel"
date: "September 2020"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::knit_hooks$set(output = function(x, options){
  if(!is.null(options$max_height)){
    paste('<pre style = "max-height:', options$max_height, '; float: left; width: 910px; overflow-y: auto;">', x, "</pre>", sep = "")
  }else{
    x
  }
  
})
```

## Data Visualisation
Data visualisation is imperative to data analysis. Representing data through a visual medium is an effective means of communicating key insights of the analysis. 

In R, we primarily use <span style="color:blue">**ggplot2**</span>, one of the most versatile and effective packages for data visualisation. To install and load this package, we can use the following code:

```{r, eval=FALSE}
# Installing Tidyverse (contains ggplot2)

install.packages("tidyverse")
library(tidyverse)

# Alternatively

install.packages("ggplot2")
library(ggplot2)
```


```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# Background -- Loading packages
library(ggplot2)
library(dplyr)
```

* *Reminders*:
  + Only need to install a package once, but necessary to reload every session. 
  + If required, we can explicitly call a specific function (from a package): `ggplot2::ggplot()` will call the `ggplot()` function from **ggplot2** package 
  
## Preliminary Steps

Think about what *visual medium* is ideal for communicating your data.

Is your data <span style="color:lightblue">Quantitative</span> or <span style="color:violet">Qualitative</span>?

If it's <span style="color:lightblue">quantitative</span>, is it:

  + Continuous or Discrete?

If it's <span style="color:violet">qualitative</span>, is it:

  + Binomial, Nominal or Ordinal?

Classifying data is important as it enables you to choose the *best visual medium* to represent your data. 

<br>

Don't know what to choose? 

* Check out:
https://www.visual-literacy.org/periodic_table/periodic_table.html# 

## The Dataset - Top Spotify Songs from 2010-2019 

* Most popular songs in the world <span style="color:green">by year</span>
* <span style="color:red">13 variables</span> to be explored
  + <span style="color:lightblue">Artist, Genre, Year, BPM, Energy, Danceability, dB, Liveness, Valence, Duration, Acousticness, Speechiness, Popularity</span>

<br> 
Download the `.csv` file <span style="color:blue">top10s.csv</span> and import it into R.

<br>
Recall:

```{r, echo=TRUE, eval=TRUE}
# Set Working Directory
setwd("C:/Users/Daniel/Desktop")

# Import dataset
top10s <- read.csv("top10s.csv")

```

## ggplot2 - Geom Functions

In this workshop, we'll cover how to implement the <span style="color:green">most common and important visualisation methods</span> that you may already be familiar with. These can be implemented by calling the function tied to the method; usually in the format of `geom_function` in <span style="color:blue">**ggplot2**</span>.

* Bar Graphs
* Histograms
* Scatter Plots
* Boxplots
* Pie Graphs
* Line Graphs
* Time Series
* Heatmaps

There are *many* other visualisation methods not included here, so if you're interested, you could check out a more extensive list:
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html

## ggplot2 - Creating a ggplot (Basics)

With ggplot2, you always begin plotting by <span style="color:violet">creating a blank, coordinate system</span> that you can then add layers to. We specify the dataset to be used for the graph, using the first argument of ggplot.

```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.width=6}
ggplot(data = top10s)
```

## ggplot2 - Mapping aesthetics in geom functions (Basics)

We complete the graph by adding <span style="color:green">*layers*</span> to `ggplot()`. This involves specifying a <span style="color:violet">geom function</span>, and mapping the variables in the dataset to visual properties. 

The mapping argument (of any geom function) is always paired with `aes()`, and the `x` and `y` arguments of `aes()` specify the variables that are mapped to the x and y axes. 

Generally, we have the following template: 

```{r, echo=TRUE, eval=FALSE}
ggplot(data = top10s) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)) 
```

*Note that we will replace* **GEOM_FUNCTION** *with a particular function, and* **MAPPINGS** *with the desired mappings.*

Also note that the **+** operator is contained within the packages, and is specific to `ggplot` objects; used to add layers to plots.

## ggplot2 - Bar Graph: General

A simple chart which diagrammatically compares the **frequency distribution** of <span style="color:violet">**discrete**</span> categorical variables.

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we examine the distribution of <span style="color:red">**Genres**</span> given by the `top.genre` variable, by using the function <span style="color:purple">**geom_bar**</span>.

```{r, echo=TRUE, eval=TRUE, fig.height=3.3, fig.width=6, fig.align='center'}
ggplot(data = top10s) +
  geom_bar(mapping = aes(x = top.genre)) 
```

## ggplot2 - Bar Graph: Changing the aesthetics

```{r, echo=TRUE, eval=TRUE, fig.height=3.7, fig.width=6, fig.align='center'}
ggplot(data = top10s) +
  geom_bar(mapping = aes(x = top.genre, fill = top.genre)) + # Colour - inside aes()
  labs(title = "Distribution of Top Genres") + # Add title
  xlab("Genre") + # x-axis label
  ylab("Frequency") + # y-axis label
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6), 
        legend.position = "none") # Labels and Adjustments
```

## ggplot2 - Histograms: General

Similar to bar graphs, but diagrammatically compares the **frequency distribution** of <span style="color:violet">**continuous**</span> variables. Histograms plot continuous data by grouping the ranges of the data into bins/intervals. 

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we examine the distribution of <span style="color:red">**BPM**</span> given by the `bpm` variable, by using the function <span style="color:purple">**geom_histogram**</span>.

```{r, echo=TRUE, eval=TRUE, fig.height=3, fig.width=6, fig.align='center', message = FALSE}
ggplot(data = top10s) +
  geom_histogram(mapping = aes(x = bpm), binwidth = 10) # Bins of width 10
```

## ggplot2 - Histograms: Changing the Aesthetics
```{r, echo=TRUE, eval=TRUE, fig.height=3.7, fig.width=6, fig.align='center'}
ggplot(data = top10s) +
  geom_histogram(mapping = aes(x = bpm), binwidth = 10, fill = "lightblue", 
                 colour = "coral") + # Coral linetype and Lightblue fill (outside aes())
  labs(title = "Distribution of BPM") + # Add title
  xlab("BPM") + # x-axis label
  ylab("Frequency") # y-axis label
```

## ggplot2 - Scatterplots: General
Used to visualise the **relationship** <span style="color:violet">**two continuous**</span> variables. <span style="color:green">Each data point is plotted</span> with x and y coordinates corresponding to the numerical value of the two variables. 

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we examine the relationship between <span style="color:red">**Speechiness**</span> and <span style="color:red">**Popularity**</span> given by `spch` and `pop` variables, by using the function <span style="color:purple">**geom_point**</span>.

```{r, echo=TRUE, eval=TRUE, fig.height=3, fig.width=6, fig.align='center', message = FALSE}
ggplot(data = top10s) +
  geom_point(mapping = aes(x = spch, y = pop)) # x - Speechiness, y - Popularity
```

## ggplot2 - Scatterplots: Changing the Aesthetics

```{r, echo=TRUE, eval=TRUE, fig.height=3.5, fig.width=6, fig.align='center', message = FALSE}
ggplot(data = top10s) +
  geom_point(mapping = aes(x = spch, y = pop, colour = top.genre)) + # Colours based on genre
  labs(title = "Speechiness vs Popularity") + # Add title
  xlab("Speechiness") + # x-axis label
  ylab("Popularity") + # y-axis label
  theme(legend.position = "none") # Remove legend
```

## ggplot2 - Boxplots: General

Representing <span style="color:violet">**groups of numerical data**</span> through quartiles. Lower and upper quartiles shown as <span style="color:green">lines extending from a rectangle</span> with most of the distribution (and the median) <span style="color:green">within the rectangle</span>.

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we compare <span style="color:red">**Popularity**</span> for **each** <span style="color:red">**Genre**</span>. Provided by the `top.genre` and `pop` variables and using the function <span style="color:purple">**geom_boxplot**</span>.

```{r, echo=TRUE, eval=TRUE, fig.height=3, fig.width=6, fig.align='center'}
ggplot(data = top10s) +
  geom_boxplot(mapping = aes(x = top.genre, y = pop)) 
```

## ggplot2 - Boxplots: Changing the Aesthetics

```{r, echo=TRUE, eval=TRUE, fig.height=3.5, fig.width=6, fig.align='center', message = FALSE}
ggplot(data = top10s) +
  geom_boxplot(mapping = aes(x = top.genre, y = pop, 
                             fill = top.genre, alpha = 0.5)) + # Colour and transparency
  labs(title = "Distribution of Popularity for each Genre") + # Add title
  xlab("Genre") + # x-axis label
  ylab("Popularity") + # y-axis label
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 3), 
        legend.position = "none") # Labels and Adjustments
```

## ggplot2 - Pie Chart: General
Circle which is <span style="color:green">divided into sectors</span> that each represent a <span style="color:violet">**proportion**</span> of the whole. 

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we examine the distribution of <span style="color:red">**Year**</span> for the **most popular Spotify songs**, provided by the `year` variable. A combination of the functions <span style="color:purple">**geom_bar**</span> and <span style="color:purple">**coord_polar**</span> is used.

```{r, echo=TRUE, eval=TRUE, fig.height=1.5, fig.width=1.5, fig.align='center'}
# Get proportions using dplyr package and create new dataframe
pie.df <- top10s %>%
  group_by(year) %>%
  summarise(n=n()) %>%
  mutate(n/sum(n)*100.0) %>%
  arrange(desc(n))

# Plot pie chart using new data frame using height of bars to represent values
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = "", y = as.factor(year)), stat = "identity", width = 1) +
  coord_polar("y", start = 0) # Making it circular
```

## ggplot2 - Pie Chart: Changing the Aesthetics

```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.width=5, fig.align='center', warning = FALSE}
ggplot(data = pie.df) + 
  geom_bar(mapping = aes(x = "", y = as.factor(year), fill = as.factor(year)), # Colour
           stat = "identity", width = 1) + 
  coord_polar("y", start = 0) +
  theme_void() + # Remove background, grid and numeric labels
  scale_fill_discrete(name = "Year") # Change legend title
```

## ggplot2 - Line Graphs (Time Series): General
Line graphs depict the <span style="color:green">relationship between two variables</span>, connecting individual data points that display <span style="color:violet">**quantitative**</span> y-values over another quantitative or qualitative x-value (usually time). **Time Series** graphs are essentially line charts, except that the horizontal axis is exclusively <span style="color:green">time-based</span>. 

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we examine how **average** <span style="color:red">**BPM**</span> changes over time (in <span style="color:red">**Years**</span>) given by `bpm` and `year` variables, by using the function <span style="color:purple">**geom_line**</span>.

*First, we'll need to average the BPMs for each Year (2010-2019) before plotting.*

```{r, echo=TRUE, eval=TRUE, max_height="100px"}
# Creating new dataframe
line.df <- top10s %>%
  group_by(year) %>% # Group by Year
  summarise(avg_bpm = mean(bpm)) # Average BPM for each Year

line.df
```

## ggplot2 - Line Graphs (Time Series): Plotting

```{r, echo=TRUE, eval=TRUE, fig.height=4.5, fig.width=6, fig.align='center'}
ggplot(data = line.df) +
  geom_line(mapping = aes(x = year, y = avg_bpm)) 
```

## ggplot2 - Line Graphs (Time Series): Aesthetics

```{r, echo=TRUE, eval=TRUE, fig.height=3.5, fig.width=6, fig.align='center'}
ggplot(data = line.df) +
  geom_line(mapping = aes(x = year, y = avg_bpm), colour = "lightblue", lwd = 1.5) +
  geom_point(aes(x = year, y = avg_bpm), colour = "coral", size = 3) +  # Overlay geom_point
  scale_x_continuous(breaks = 2010:2019) + # Changing ticks
  labs(title = "Average BPM from 2010-2019") + # Add title
  xlab("Year") + # x-axis label
  ylab("Average BPM") # y-axis label
```

## Base R - Heatmaps: General

Heatmaps represent the <span style="color:green">magnitude</span> of <span style="color:violet">**numerical**</span> values by **colours**, providing an immediate visual summary of the data. Heatmaps are versatile; able to adapt to various forms of analytics. 

<hr/>

In our dataset <span style="color:blue">**top10s**</span>, we will construct a **correlation** heatmap between <span style="color:red">**all 10 numeric variables**</span> (excluding Year), by using the <span style="color:purple">**heatmap()**</span> function in <span style="color:blue">**Base R**</span>.

*First, we need to output a correlation matrix for our 10 variables.*

```{r, echo=TRUE, eval=TRUE, max_height="130px"}
corrmat <- round(cor(top10s[,6:15]), 2) # Correlation Matrix of last 10 Variables, rounded

corrmat # Display Correlation Matrix
```

## Base R - Heatmaps: Plotting

```{r, echo=TRUE, eval=TRUE, fig.align='center', fig.height = 5}
heatmap(corrmat)
```

## Base R - Heatmaps: Aesthetics

```{r, echo=TRUE, eval=TRUE, fig.align='center', fig.height = 5}
# Base R's colour palettes: terrain.color(), rainbow(), heat.colors(), topo.colors()
heatmap(corrmat, Colv = NA, Rowv = NA, col = cm.colors(256), main = "Correlation Heatmap")
```

## ggplot2 - Bonus: Overlaying Graphs 1

* Overlaying a <span style="color:purple">**Scatter Plot**</span> and a <span style="color:purple">**Smooth Curve OBF**</span>. 
```{r, echo=TRUE, eval=TRUE, fig.align='center', fig.height = 4, message = FALSE}
ggplot(top10s, aes(x = spch, y = pop, colour = bpm, alpha = 0.7)) +
  geom_point() +
  geom_smooth()
```

## ggplot2 - Bonus: Overlaying Graphs 2
* Overlaying <span style="color:purple">**two Histograms**</span>.

```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.align='center'}
ggplot(top10s) + 
  geom_histogram(aes(x = nrgy), binwidth = 10, fill = "lightblue", alpha = 0.3) +
  geom_histogram(aes(x = dnce), binwidth = 10, fill = "lightpink", alpha = 0.3)
```

## ggplot2 - Bonus: Allocating sizes

* Sizes of circles in  <span style="color:purple">**Scatterplot**</span> are indicative of the **magnitude** of another variable. 
```{r, echo=TRUE, eval=TRUE, fig.align='center', fig.height = 4}
ggplot(top10s) +
  geom_point(aes(x = spch, y = pop, colour = top.genre, size = bpm), alpha = 0.3) + # Scale circle size depending on BPM
    theme(legend.position = "none") # Remove legend
```

## ggplot2 - Bonus: Stacked bar graphs

* Mapping the <span style="color:coral">**fill**</span> aesthetic to another variable `artist`: Bars become stacked and each rectangle (representing a certain `genre`) comprises of the `artists` that make up the `genre`. 
```{r, echo=TRUE, eval=TRUE, fig.height=3.8, fig.align='center'}
ggplot(data = top10s) +
  geom_bar(mapping = aes(x = top.genre, fill = artist), alpha = 0.7) + # Fill according to artist
  xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6), legend.position = "none")
```

## Extension - Defining functions for efficiency

Functions can be used to prevent repetition when setting the parameters, especially when multiple, similar plots are required.

```{r, echo=TRUE, eval=TRUE}
create_scatter <- function(df, xvar, yvar, xname, yname, title){
  ggplot(df) +
    geom_point(aes(x = {{xvar}}, y = {{yvar}}, colour = pop)) + # Colour based on Popularity
    labs(title = title) + # Title
    scale_x_discrete(name = xname) + # x-axis label
    scale_y_continuous(name = yname) + # y-axis label
    theme(legend.position = "none") # Remove legend
}
```

<span style="color:red">**Example**</span>

```{r, echo = TRUE, eval = FALSE}
# Create a scatterplot (using user-defined function) of BPM and Liveliness
create_scatter(top10s, bpm, live, "Beats Per Minute", "Liveliness", "BPM vs Liveliness")
```

Note: *Use `{{` and `}}` when defining which variable to call when creating your function, otherwise it will return* `object 'variable' not found`. 

## Extension - Automating Plotting Process
We can use <span style="color:orange">**loops**</span> function to create plots *quickly and easily*. First, we create an empty list:

```{r, echo = TRUE, eval = FALSE}
plot_list = list() # Empty list
```

Then, we create a <span style="color:orange">**for**</span> loop to store the *ith variable plot* (with `popularity` variable) in the list, iterated:
```{r, echo = TRUE, eval = FALSE}
for(i in 6:15){ 
  plot <- ggplot(top10s) + # For each iteration, store the plot made with the ith variable and popularity
    geom_point(aes(x = top10s[,i], y = pop, colour = top.genre)) + # Scatterplot
    theme(legend.position = "none") + # Remove legend
    xlab(colnames(top10s)[i]) + # x-axis label
    ylab("Popularity") # y-axis label
  plot_list[[i]] = plot # Add the ith plot to the plot_list
} 
```

Finally, we <span style="color:orange">**export**</span> our graphs using R's native functions:
```{r, echo = TRUE, eval = FALSE}
pdf("myplots.pdf") # Create empty pdf file named "myplots"
for(i in 6:15){ # Print each plot in plot_list variable (on separate page in pdf)
  print(plot_list[[i]])
} 
dev.off() # Complete action and save PDF file
```

## ggplot2 - Exporting your graphs: `ggsave()`

`ggsave()` is a convenient function in the <span style="color:blue">**ggplot2**</span> package which defaults to saving the <span style="color:orange">**last plot displayed**</span>.

```{r, echo = TRUE, eval = FALSE}
ggsave("myfilename.png", width = 28, height = 9, dpi = 300)
# Specifying Name, Type, Width, Height and Resolution
```

* There are other arguments it can take as well:
  + `plot`: Specify which plot (otherwise it will default to last plot)
  + `device`: Device to use: e.g. "pdf", "jpeg", "png" etc. 
  + `units`: ("in","cm" or "mm")
  + `limitsize`: TRUE (default) will not save images larger than 50x50 inches

<br/>
<br/>
<br/>

See more: https://ggplot2.tidyverse.org/reference/ggsave.html 

## R - Exporting your graphs using Base R

1. After creating your graph, you can simply go to the <span style="color:violet">**Plots Panel -> Export -> Save as Image or Save as PDF**</span> and specify your dimensions and format.

2. Using code:

* First, specify the file format.
* Enter your plotting commands (or call the saved plot)
* Then use `dev.off()` command.

```{r, eval = FALSE, echo = TRUE}
jpeg("my_plot_name.jpg") # Specify format and name
plot(x,y) OR myplot # Plot the graph or call the graph
dev.off() # To complete process (important)
```

Common Formats:

* `pdf("rplot.pdf")`: PDF file
* `png("rplot.png")`: PNG file
* `jpeg("rplot.jpg")`: JPEG file
* `bmp("rplot.bmp")`: BMP file

## Useful links for Data Visualisation in R

* ggplot2 Cheatsheet
  + https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf 

<br/>

* Base R - Data Visualisation Cheatsheet
  + https://www.analyticsvidhya.com/wp-content/uploads/2015/08/Comprehensive-Guide-to-Data-Visualization-in-R-FINAL-1.jpg 

<br/>

* Periodic Table of Visualisation Methods
  + https://www.visual-literacy.org/periodic_table/periodic_table.html# 

<br/>

* More Visualisation Methods in R
  + http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html 
  
